---
title: "Infer ASVs with DADA2"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center", 
                      fig.path = "../figures/01_DADA2/")

```

```{r set_seed, include=FALSE}
# Set my seed - can choose any number
set.seed(23598)
```

# Goals of this file
1. Use raw fastq files and generate quality plots to assess quality of reads.
2. Filter and trim out bad seqs and bases from seq files.
3. Write out fastq files with high quality seqs. 
4. Evaluate quality from filter and trim. 

# Load libraries
```{r load_libraries}
#install.packages("devtools")
library(devtools)

#devtools::install_github("benjjneb/dada2")
library(dada2)
```

# Load data
```{r load_data}
# Set raw fastq path to the raw seq files
# PATH to fastq files

raw_fastq_path <- "data/01_DADA2/01_raw_gzip_fastqs"

# What files are in this path? 
head(raw_fastq_path)

# How many files are there?
str(list.files(raw_fastq_path))

# Create a vector of forwards reads
forward_reads <- list.files(raw_fastq_path, pattern = "R1_001.fastq.gz", full.names = TRUE)
head(forward_reads)

# Create a vector of reverse reads
reverse_reads <- list.files(raw_fastq_path, pattern = "R2_001.fastq.gz", full.names = TRUE)
head(reverse_reads)
```

# Quality plots
```{r raw_quality_plot}
# Randomly select two samples from dataset to evaluate
random_samples <- sample(c(1:length(reverse_reads)),2)

# Calculate and plot the quality of two samples
plotQualityProfile(forward_reads[random_samples])
plotQualityProfile(reverse_reads[random_samples])
```

# Prepare placeholder for filtered reads
```{r prep_filtered_seqs}
# Make vector of our samples, extract sample name from files
samples <- sapply(strsplit(basename(forward_reads), "_"),`[`,1)
head(samples)

# Place filtered reads into filtered fastq paths
filt_fastq_path <- "/local/workdir/rf384/git_repos/SalinityGradient16S/data/01_DADA2/02_filtered_fastqs"

# Create 2 vars - one for forward, one for reverse
filt_forward_reads <- file.path(filt_fastq_path, paste0(samples, "_R1_filtered.fastq.gz"))
filt_reverse_reads <- file.path(filt_fastq_path, paste0(samples, "_R2_filtered.fastq.gz"))
```


# Filter and trim reads
Parameters of filter and trim depend on the dataset
- maxN: number of N bases. Default is $0$, removes all Ns from dataset.
- maxEE: quality filtering threshold applied to expected errrors. Default is $2$, allows up to $2$ expected errors. There are two values - the first value is for forward reads, the second for reverse.  
- trimLeft: Number of bases to remove from beginning of the read (here $3$). 
- trunQ: Truncate reads at the first instance of a quality score less than or equal to truncQ.

```{r filter_and_trim}
# Assign a vector to filtered reads
# Trim out poor bases and first 3 bps on F reads
# Write out fastq files
filtered_reads <- filterAndTrim(fwd = forward_reads, filt = filt_forward_reads, 
                     rev = reverse_reads, filt.rev = filt_reverse_reads,
                     maxEE = c(2,2), trimLeft = 3, truncQ = 2, rm.phix = TRUE,
                     compress = TRUE) #multithread = TRUE
```



# Session information
```{r session_info}
# Ensure reproducibility
devtools::session_info()
```

