---
title: "Phylogenetic Tree Inspection & Rooting"
output: 
  html_document:
    toc: yes
    toc_float: yes
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center", 
                      fig.path = "../figures/03_Phylogenetic_Tree/")

```

# Goals

1. Load the fasttree unrooted tree.
2. Add tree to phyloseq object.
3. Visualize and inspect tree with ggtree.
4. Prune ASVs if needed.
5. Combine new tree with phyloseq object.
6. Save 2 phyloseq objects: 1. Unrooted phyloseq object, 2. Rooted phyloseq object

# Set the environment
## Set seed
```{r set_seed, include=FALSE}
# Set my seed - can choose any number
set.seed(23598)
```

## Load packages
```{r load_packages}
pacman::p_load(tidyverse, phyloseq, ggtree, phytools, install=FALSE)
```

## Load data
```{r load_data}
# Load phyloseq object
load("data/02_Phyloseq/raw_preprocessed_physeq.RData")
row_preprocessed_physeq

# Load in tree
unrooted_tree <- read.tree("data/03a_Phylogenetic_Tree/ASVs_unrooted.tree")
unrooted_tree
str(unrooted_tree)
```

# Merge Phyloseq
```{r merge_physeq}
# Intuition check
stopifnot(ntaxa(row_preprocessed_physeq) == ntaxa(unrooted_tree)) 

# Merge tree with phyloseq
unrooted_physeq <- merge_phyloseq(row_preprocessed_physeq, unrooted_tree)
```

# Plot tree with ggtree
```{r plot_tree_unrooted}
# Make a basic tree
kingdom_node_tree <- ggtree(unrooted_physeq) + 
  # color by kingdom
  geom_tippoint(mapping = aes(color = Kingdom)) +
  scale_color_manual(values = c("goldenrod1", "cornflowerblue", "gray")) +
  labs(title = "Unrooted Tree") +
  # add node label
  geom_text(aes(label = node), hjust = -0.5, vjust = -0.5, size = 2) +
  theme(legend.position = "bottom") 
```

## Evaluate long branch
```{r eval_long_branch}
# View clade 2537
viewClade(kingdom_node_tree, node = 2537) +
  labs(title = "Unrooted Tree: Node 2537") 

# Look into clade
viewClade(kingdom_node_tree, node = 5208) +
  labs(title = "Unrooted Tree: Node 2537") + 
  geom_text(aes(label = ASV))

# Pull out information about ASV
unrooted_physeq %>%
  subset_taxa(., ASV == "ASV_456") %>%
  tax_table %>%
  data.frame

```


# Session information

