---
title: "Phylogenetic Tree Inspection & Rooting"
output: 
  html_document:
    toc: yes
    toc_float: yes
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center", 
                      fig.path = "../figures/03_Phylogenetic_Tree/")

```

# Goals

1. Load the fasttree unrooted tree.
2. Add tree to phyloseq object.
3. Visualize and inspect tree with ggtree.
4. Prune ASVs if needed.
5. Combine new tree with phyloseq object.
6. Save 2 phyloseq objects: 1. Unrooted phyloseq object, 2. Rooted phyloseq object

# Set the environment
## Set seed
```{r set_seed, include=FALSE}
# Set my seed - can choose any number
set.seed(23598)
```

## Load packages
```{r load_packages}
pacman::p_load(tidyverse, phyloseq, ggtree, phytools, install=FALSE)
```

## Load data
```{r load_data}
# Load phyloseq object
load("data/02_Phyloseq/raw_preprocessed_physeq.RData")
row_preprocessed_physeq

# Load in tree
unrooted_tree <- read.tree("data/03a_Phylogenetic_Tree/ASVs_unrooted.tree")
unrooted_tree
#str(unrooted_tree)
```

# Merge Phyloseq
```{r merge_physeq}
# Intuition check
stopifnot(ntaxa(row_preprocessed_physeq) == ntaxa(unrooted_tree)) 

# Merge tree with phyloseq
unrooted_physeq <- merge_phyloseq(row_preprocessed_physeq, unrooted_tree)
```

# Plot tree with ggtree
```{r plot_tree_unrooted}
# Make a basic tree
kingdom_node_tree <- ggtree(unrooted_physeq) + 
  # color by kingdom
  geom_tippoint(mapping = aes(color = Kingdom)) +
  scale_color_manual(values = c("goldenrod1", "cornflowerblue", "gray")) +
  labs(title = "Unrooted Tree") +
  # add node label
  geom_text(aes(label = node), hjust = -0.5, vjust = -0.5, size = 2) +
  theme(legend.position = "bottom") 

kingdom_node_tree
```

## Evaluate long branch
```{r eval_long_branch}
# View clade 2537
viewClade(kingdom_node_tree, node = 2537) +
  labs(title = "Unrooted Tree: Node 2537") 

# Look into clade
viewClade(kingdom_node_tree, node = 5208) +
  labs(title = "Unrooted Tree: Node 2537") + 
  geom_text(aes(label = ASV))

# Pull out information about ASV
unrooted_physeq %>%
  subset_taxa(., ASV == "ASV_456") %>%
  tax_table %>%
  data.frame

```
Alphaproteobacteria with few counts - mitochondria associated.

# Remove ASV 456
```{r remove_ASV_456}
# Function from Joey McMurdie: https://github.com/joey711/phyloseq/issues/652
pop_taxa = function(physeq, badTaxa){
  allTaxa <-  taxa_names(physeq)
  allTaxa <- allTaxa[!(allTaxa %in% badTaxa)]
  return(prune_taxa(allTaxa, physeq))}

# Use pop taxa function
# Recreate phyloseq object w/o ASV_456
unrooted_physeq_rm456 <- unrooted_physeq %>%
  pop_taxa(., "ASV_456")

# Intuition check
ntaxa(unrooted_physeq) - ntaxa(unrooted_physeq_rm456)

# Plot it
ggtree(unrooted_physeq_rm456) + 
  geom_tippoint(mapping = aes(color = Kingdom))
```

# Midroot tree
```{r midroot_tree}
# Is the tree rooted?
new_unrooted_tree <- phy_tree(unrooted_physeq_rm456)
is.rooted(new_unrooted_tree)


# Root the tree
midpoint_rooted_tree <- midpoint.root(new_unrooted_tree)
is.rooted(midpoint_rooted_tree)

# Assign to a new phyloseq object: merging subsetted phyloseq with new rooted tree
# Create phyloseq object w/o ASV 456
physeq_rm456 <- row_preprocessed_physeq %>%
  subset_taxa(ASV != "ASV_456")

# Merge tree with new physeq
midroot_physeq_rm456 <- merge_phyloseq(physeq_rm456, midpoint_rooted_tree)
midroot_physeq_rm456

# Inspect tree
ggtree(midroot_physeq_rm456) + 
  geom_tippoint(mapping = aes(color = Kingdom))
```

# Export physeq objects
```{r save_physeq}
save(list = c("unrooted_physeq_rm456", "midroot_physeq_rm456"), 
     file = "data/03a_Phylogenetic_Tree/Phytree_Preprocessed_Physeq.RData")
```


# Session information
```{r session_info}
devtools::session_info()